---
title: "Data exploration of immune data in laboratory and field infections"
author: "Fay"
date: "2024-02-01"
output: 
    html_document:
        toc: yes # table of content
        toc_float: 
            collapsed: false # show full toc
            smooth_scroll: true # toc scrolling behavior
    pdf_document:
         keep_md: true
         fig_width: 12
         fig_height: 8
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)

```

```{r include = FALSE, message = FALSE, warning=FALSE}
# Libraries
library(dplyr)
library(tidyr)
library(ggplot2)
library(ggridges)
library(knitr)
```

```{r include = FALSE, message = FALSE, warning=FALSE}
# Read data
hm <- read.csv("Data/Data_output/imputed_clean_data.csv")

# divide into laboratory and field data set
lab <- hm %>% filter(origin == "Lab")
field <- hm %>% filter(origin == "Field")

###################### eimeria strains
# define the color for each level of infection
color_mapping <- c("E_falciformis" = "salmon", 
                   "E_ferrisi" = "forestgreen", 
                   "uninfected" = "blue")

```


# Data exploration of field and laboratory infection data 
## Laboratory data compilation of experimental infections 
### Experimental planning 
#### Mice 
* four wild-derived inbred mouse strains along with their respective F1 hybrids. 
* Two of these strains, SCHUNT and STRA, represent M. m. domesticus. 
* The strains BUSNA and PWD were derived from M. m. musculus
* Two intersubspecific hybrids (STRAxBUSNA and SCHUNTxPWD) and 
* Two intrasubspecific hybrids (SCHUNTxSTRA and PWDxBUSNA). 

```{r mouse_strains, warning=FALSE, echo = FALSE}
mouse_strains <- c("SCHUNT", "STRA", "BUSNA", "PWD", "STRA BUSNA", 
                   "SCHUNT PWD", "SCHUNT STRA", "PWD BUSNA")
hybrid_status <- c("M. m. domesticus", "M. m. domesticus", "M. m. musculus",
                   "M. m. musculus", "intersubspecific hybrids",
                   "intersubspecific hybrids", "intrasubspecific hybrids", 
                   "intrasubspecific hybrids")
strains <- as.data.frame(hybrid_status, mouse_strains)
kable(strains)
```


```{r strains,  warnings = FALSE}
lab %>%
    group_by(mouse_strain) %>%
    # Summarize the data to get counts for each mouse strain
    summarize(count = n()) %>%
    # Reorder mouse_strain by count
    mutate(mouse_strain = reorder(mouse_strain, count)) %>%
    # Plotting
    ggplot(aes(x = mouse_strain, y = count, fill = mouse_strain)) +
    geom_bar(stat = "identity") + 
    # Specify stat = "identity" for pre-summarized data
    geom_text(aes(label = count), vjust = -0.3) + 
    # Add count labels on top of bars
    scale_fill_viridis_d() + 
    # Use a nice color scale, like Viridis
    theme_minimal() + # Apply a minimal theme for a cleaner look
    theme(axis.text.x = element_text(angle = 50)) +
    labs(title = "Mouse Strain in experimental infections", x = "Mouse Strain", 
         y = "Count") +# Add label
    guides(fill = FALSE)
```


```{r}
# Define colors
colors <- c("TRUE" = "firebrick3", "FALSE" = "steelblue")

# transform mouse strain into factor
lab$mouse_strain <- as.factor(lab$mouse_strain)

lab$mouse_strain <- gsub(pattern = "_", " ", lab$mouse_strain)

# order factor levels
lab$mouse_strain <- factor(lab$mouse_strain, 
                           levels = names(
                               sort(tapply(lab$WL_max, lab$mouse_strain, median))))

ggplot(lab %>%
           filter(infection == "challenge"), 
       aes(x = WL_max, y = mouse_strain, fill = mouse_strain)) + 
    geom_density_ridges(jittered_points = TRUE, position = position_points_jitter(height = 0), 
                        scale = 0.9, alpha = 0.6, point_shape = 21, point_size = 2, point_alpha = 1) +
    geom_boxplot(width = 0.2, outlier.shape = NA, alpha = 0.5, position = position_nudge(x = 0.2)) +
    coord_flip() +
    theme_minimal() +
    theme(legend.position = "none",
          axis.text.x = element_text(angle = 70, vjust = 0.5, hjust=0.3)) +
    xlab("Maximum relative weight loss") +
    ylab("Mouse Strain") -> strains_weight

strains_weight

ggsave(filename = "figures/strains_weight.jpeg", plot = strains_weight, 
       width = 8, height = 6, dpi = 1000)

ggplot(lab, aes(x = WL_max, y = mouse_strain, fill = mouse_strain)) + 
    geom_density_ridges(jittered_points = TRUE, position = position_points_jitter(height = 0), 
                        scale = 0.9, alpha = 0.6, point_shape = 21, point_size = 2, point_alpha = 1) +
    geom_boxplot(width = 0.2, outlier.shape = NA, alpha = 0.5, position = position_nudge(x = 0.2)) +
    coord_flip() +
    theme_minimal() +
    theme(legend.position = "none",
          axis.text.x = element_text(angle = 70, vjust = 0.5, hjust=0.3)) +
    xlab("Maximum relative weight loss") +
    ylab("Mouse Strain") +
    facet_grid(~infection) -> strains_weight_challenge

strains_weight_challenge

ggsave(filename = "figures/strains_weight_chalenge.jpeg", plot = strains_weight_challenge, 
       width = 10, height = 6, dpi = 1000)

###################### eimeria strains
# Then, define the color for each level of infection
color_mapping <- c("E_falciformis" = "salmon", 
                   "E_ferrisi" = "forestgreen", 
                   "uninfected" = "blue")

lab$current_infection <- gsub(pattern = "_", replacement = ". ", lab$current_infection)

ggplot(lab %>%
           filter(infection == "challenge"), 
       aes(x = WL_max, y = current_infection, fill = current_infection)) + 
    geom_density_ridges(jittered_points = TRUE, position = position_points_jitter(height = 0), 
                        scale = 0.9, alpha = 0.6, point_shape = 21, point_size = 2, point_alpha = 1) +
    geom_boxplot(width = 0.2, outlier.shape = NA, alpha = 0.5, position = position_nudge(x = 0.2)) +
   # coord_flip() +
    theme_minimal() +
    scale_color_manual(values = color_mapping) +
    theme(legend.position = "none",
          axis.text.x = element_text(angle = 0, vjust = 0.5, hjust=0.3)) +
    xlab("Maximum relative weight loss") +
    ylab("Parasite strain") -> eimeria_weight

eimeria_weight

ggsave(filename = "figures/eimeria_strains_weight.jpeg", plot = eimeria_weight, 
       width = 8, height = 6, dpi = 1000)

# prim vs challenge
lab$current_infection <- gsub(pattern = "_", replacement = ". ", lab$current_infection)

ggplot(lab, 
       aes(x = WL_max, y = current_infection, fill = current_infection)) + 
    geom_density_ridges(jittered_points = TRUE, position = position_points_jitter(height = 0), 
                        scale = 0.9, alpha = 0.6, point_shape = 21, point_size = 2, point_alpha = 1) +
    geom_boxplot(width = 0.2, outlier.shape = NA, alpha = 0.5, position = position_nudge(x = 0.2)) +
    # coord_flip() +
    theme_minimal() +
    theme(legend.position = "none",
          axis.text.x = element_text(angle = 0, vjust = 0.5, hjust=0.3)) +
    xlab("Maximum relative weight loss") +
    ylab("Parasite strain") +
    scale_color_manual(values = color_mapping) +
    facet_wrap(~ infection, nrow = 2)-> eimeria_weight_challenge

eimeria_weight_challenge

ggsave(filename = "figures/eimeria_strains_weight_challenge.jpeg", plot = eimeria_weight_challenge, 
       width = 8, height = 10, dpi = 1000)


###### is the difference between parasite strains significant?
parasite_WL <- lm(formula = WL_max ~ current_infection, data = lab)
summary(parasite_WL)


##### is the differnece between mouse strains significant?`
mouse_WL <- lm(WL_max ~ mouse_strain, data = lab)
summary(mouse_WL)               


```



