---
title: "Norm_Imputation_strategy_selection"
author: "Fay"
date: "2023-09-28"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Normalize or impute first? 

Here is a document on selecting the best order for normalizationa and imputation of the immune gene expression data. 

## Layout: 

1. Correlation of non-normalized and non-imputed gene expression data
2. Correlation of non-normalized and imputed gene expression data 
3. Correlation of normalized data (no imputation)
4. Correlation of 1st normalized and sequentially imputed gene expression data
5. Corrrelation of 1st imputed and sequentially normalized gene expression data

## Data input

### Libraries
```{r, warning=FALSE, message=FALSE, include = FALSE}
library(mice)
library(stringr)
library(gridExtra)
library(dplyr)
library(tidyverse)
library(tidyr)
library(janitor)
library(visdat)
library(corrplot)
library(RColorBrewer)
library(ggplot2)
library(VIM)


```


```{r, warning=FALSE, message=FALSE, include = FALSE}
#Lab genes
# The measurements of IL.12 and IRG6 are done with an other assay and will 
#ignore for now
Genes_v   <- c("IFNy", "CXCR3", "IL.6", "IL.13", "IL.10",
                "IL1RN","CASP1", "CXCL9", "IDO1", "IRGM1", "MPO", 
                "MUC2", "MUC5AC", "MYD88", "NCR1", "PRF1", "RETNLB", "SOCS1", 
                "TICAM1", "TNF") #"IL.12", "IRG6")
```


```{r, warning=FALSE, message=FALSE, include = FALSE}
hm <- read.csv("Data/Data_output/1.MICE_cleaned_data.csv")


hm$Mouse_ID <- str_replace(hm$Mouse_ID, "_", "")

field <- hm %>%
    dplyr::filter(origin == "Field") 

# select the genes, the mouse identifier and the house keeping gene
gmf <- field[, c("Mouse_ID", Genes_v, "GAPDH")] 

# Remove columns with only NA values
gmf <- gmf %>% select_if(~!all(is.na(.)))

#remove rows with only nas 
gmf <- gmf[!apply(is.na(gmf[-1]), 1, all), ]

gf <- gmf[,-1]

##select same rows in the first table
field <- field %>% 
    filter(Mouse_ID %in% gmf$Mouse_ID)


###############lab
#select the genes and lab mice
lab <- hm %>%
    dplyr::filter(origin == "Lab", Position == "mLN") %>%
    group_by(Mouse_ID, infection) %>%
    filter(dpi == max_dpi)
    

# select the genes, the mouse identifier and the house keeping gene
gml <- lab[, c("Mouse_ID", Genes_v, "PPIB")] 

gml <- unique(gml)

# Remove columns with only NA values
gml <- gml %>% select_if(~!all(is.na(.)))

#remove rows with only nas 
gml <- gml[!apply(is.na(gml[-1]), 1, all), ]

gl <- gml[,-1]

##select same rows in the first table
lab <- lab %>%
    filter(Mouse_ID %in% gml$Mouse_ID)


# looking at patterns of nas)
#pattern_na <-as.data.frame(md.pattern(field_genes))
#field 
sapply(gmf, function(x) sum(is.na(x)))

#lab
sapply(gml, function(x) sum(is.na(x)))

#remove duplicates
lab_prim <- lab %>%
    filter(death == "primary")
lab_chal <- lab %>% 
    filter(death == "challenge", infection == "challenge")

lab <- rbind(lab_prim, lab_chal)
rm(lab_prim, lab_chal)

hm <- rbind(lab, field)
```

## 1. Correlation of non-normalized and non-imputed gene expression data
```{r, warning=FALSE, message=FALSE, include = TRUE}
gene_correlation <- lab %>% 
  filter(infection == "challenge", dpi == max_dpi) %>%
  ungroup() %>%
  dplyr::select(all_of(Genes_v))

# draw correlation between the genes
gene_correlation <- as.matrix(cor(gene_correlation, 
                                  use="pairwise.complete.obs"))

# load the function to calculate the p value for correlations
source("R/Functions/p_value_for_correlations.R")

# matrix of the p-value of the correlatio
p.mat <- cor.mtest(gene_correlation)

corrplot(gene_correlation, 
         method = "circle",  #method of the plot, "color" would show colour gradient
         tl.col = "black", tl.srt=45, #colour of labels and rotation
         col = brewer.pal(n = 8, name ="RdYlBu"), #colour of matrix
         order="hclust", #hclust reordering
         p.mat = p.mat, sig.level = 0.01, insig = "blank",
         addCoef.col = 'black',
         number.cex=0.5, 
         title = "Lab") 
  #Add significance level to the correlogram
#remove the values that are insignificant

```

### Correlation of non-normalized and non-imputed gene expression data, Results: 
a. positive correlations between MUC2, IRGM1 and SOCS1
b. positive correlations between CASP1 and MUC5AC
c. positive correlations between IFNy, IL.10, IL.13, TNF, CXCL9, TICAM1 and IDO1
d. MYD88, PRF1, IL.6, NCR1, RETNLB, TICAM1 positve correlations with group c
e. negative correlations between group a (MUC2, IRGM1, SOCS1) and group c (IFNY, IL.10 etc)
f. negative correlations between IL1RN, MPO and group a
g. negative correlations between 
NCR1, RETNLB, TICAM1 and group a


## 2. Correlation of non-normalized and imputed gene expression data 
```{r warning=FALSE, message=FALSE, include = FALSE}
hm_genes <- hm[,c("Mouse_ID", Genes_v)]

#pattern_na <-as.data.frame(md.pattern(field_genes))
sapply(hm_genes, function(x) sum(is.na(x)))

genes <- hm_genes[, -1]

# The frequency distribution of the missing cases per variable can be obtained 
# as:
init <- mice(genes, maxit = 0)

#we want to impute only the specific variables
meth <- init$method

# removing il 10
genes <- genes[, !(names(genes) %in% "IL.10")]

# removed(because of large missing numbers)
# m=5 refers to the number of imputed datasets. Five is the default value.
igf <- mice(genes, m = 5, seed = 500) # method = meth,

summary(igf)

## igf$imp$IFNy
#Now we can get back the completed dataset using the complete()
complete_genes <- complete(igf, 1)

# remove IL.10 from the selection vector
Genes_v <- Genes_v[!Genes_v == "IL.10"]

#remove the non imputed genes from our data set
hm <-hm %>%
  dplyr::select(-all_of(Genes_v))

# join the imputed genes
Mouse_ID <- hm_genes$Mouse_ID
result <- data.frame(Mouse_ID, complete_genes)

hm_imp <- hm %>%
  left_join(result, by = "Mouse_ID")
```

```{r, warning=FALSE, message=FALSE, include = TRUE}
lab <- hm_imp %>%
    filter(origin== "Lab")
gene_correlation <- lab[,Genes_v]

# draw correlation between the genes
gene_correlation <- as.matrix(cor(gene_correlation, 
                                  use="pairwise.complete.obs"))

# matrix of the p-value of the correlatio
p.mat <- cor.mtest(gene_correlation)

corrplot(gene_correlation, 
         method = "circle",  #method of the plot, "color" would show colour gradient
         tl.col = "black", tl.srt=45, #colour of labels and rotation
         col = brewer.pal(n = 8, name ="RdYlBu"), #colour of matrix
         order="hclust", #hclust reordering
         p.mat = p.mat, sig.level = 0.01, insig = "blank",
         addCoef.col = 'black',
         number.cex=0.5,
         title = "Lab") 
  #Add significance level to the correlogram
#remove the values that are insignificant
```
### 2. Correlation of non-normalized and imputed gene expression data, Results:
a. positive correlations between MUC2, IRGM1, SOCS1
b. positive correlations between IL1RN, MPO
c. positve correlations between CASP1, MUC5AC
d. positve correlations between CXCR3, MYD88
e. positive correlations between RETNLB, TICAM1, IL.6, TNF, CXCL9, IFNy, IDO1
f. positive correlations between IL1RN and TNF, IFN< and IDO1
g. negative correlations between IRGM1, SOCS2 and IL1RN, MPO
h. negative correlations between group MUC2, IRGM1, SOC1 vs IL.6, TNF, CXCL9, IFNy and IDO1
i. negative correlations between IL.13 and CASP1, MUC5AC and MYD88

## 3. Correlation of normalized data (no imputation)
```{r warning=FALSE, message=FALSE, include = FALSE}
#ΔCt (sample) = Ct(gene of interest) − Ct(reference gene)

df <- hm %>%
    filter(origin == "field")

#check the distributions of genes
plot_list <- lapply(names(gf), function(colname){
  ggplot(df, aes_string(x = colname)) +
    geom_histogram(bins = 30, fill = "skyblue", color = "black") +
    labs(title = paste("Histogram of", colname))
})


#grid.arrange(grobs = plot_list, ncol = 3) 
#placed in hashtags, takes to long to process
calculate_expression <- function(df) {
  # Extract GAPDH column
  GAPDH <- df$GAPDH
  
  # Step 1: Calculate delta ct for all genes except GAPDH
  delta_ct <- df[, sapply(df, is.numeric)]  # Ensure only numeric columns are taken
  delta_ct <- delta_ct[, !colnames(delta_ct) %in% "GAPDH"]  # Exclude GAPDH
  delta_ct <- sapply(delta_ct, function(gene) gene - GAPDH)
  
  # Step 2: Calculate ΔΔct
  mean_dct <- colMeans(delta_ct, na.rm = TRUE)  # Calculate mean delta_ct for each gene
  delta_delta_ct <- t(apply(delta_ct, 1, function(row) row - mean_dct))
  
  # Step 3: Calculate the result
  result <- 2^(-delta_delta_ct)
  result <- round(result, 2)
  return(result)
}

# Use the function
result_df <- calculate_expression(df)

# View the result
print(result_df)

#check the distributions of genes
plot_list <- lapply(names(result_field[,-1]), function(colname){
  ggplot(result_field[,-1], aes_string(x = colname)) +
    geom_histogram(bins = 30, fill = "skyblue", color = "black") +
    labs(title = paste("Histogram of", colname))
})


#grid.arrange(grobs = plot_list, ncol = 3)

############### same for lab ##################################################
# In the lab we use PPIB
#hist(lab$PPIB)

#check the distributions of genes
plot_list <- lapply(names(gl), function(colname){
  ggplot(gl, aes_string(x = colname)) +
    geom_histogram(bins = 30, fill = "skyblue", color = "black") +
    labs(title = paste("Histogram of", colname))
})


#grid.arrange(grobs = plot_list, ncol = 3)


calculate_expression <- function(df) {
  # Extract PPIB column
  PPIB <- df$PPIB
  
  # Step 1: Calculate delta ct for all genes except PPIB
  delta_ct <- df[, sapply(df, is.numeric)]  # Ensure only numeric columns are taken
  delta_ct <- delta_ct[, !colnames(delta_ct) %in% "PPIB"]  # Exclude PPIB
  delta_ct <- sapply(delta_ct, function(gene) gene - PPIB)
  
  # Step 2: Calculate ΔΔct
  mean_dct <- colMeans(delta_ct, na.rm = TRUE)  # Calculate mean delta_ct for each gene
  delta_delta_ct <- t(apply(delta_ct, 1, function(row) row - mean_dct))
  
  # Step 3: Calculate the result
  result <- 2^(-delta_delta_ct)
  
  return(as.data.frame(result))
}

# Use the function
result_lab <- calculate_expression(gml)

# View the result
#print(result_lab)
#summary(result_lab)

#check the distributions of genes
plot_list <- lapply(names(result_lab[,-1]), function(colname){
  ggplot(result_lab[,-1], aes_string(x = colname)) +
    geom_histogram(bins = 30, fill = "skyblue", color = "black") +
    labs(title = paste("Histogram of", colname))
})


grid.arrange(grobs = plot_list, ncol = 3)



#join the normalised data
result <- rbind(result_field, result_lab)


# When working with DDct Values
# A ddct of 0 = no change in fold expression
# negative ddct = upregulation
# postive ddct = downregulation
# Check if each column is numeric and, if so, multiply by -1
#result[] <- lapply(result, function(x) if(is.numeric(x)) -1 * x else x)

# Therefore deciding to transform data by *-1
# then positive = upgregulation - more intuitive for interpretation
rm(gmf,data, gf, gl, gml, hm, result_field, result_lab)

hm <- rbind(field, lab)

#remove non normalised genes
hm <-hm %>%
  dplyr::select(-all_of(Genes_v))

# join the normalised genes
hm <- hm %>%
  left_join(result, by = "Mouse_ID")

rm(lab, field, result)

## 4. Correlation of 1st normalized and sequentially imputed gene expression data
## 5. Corrrelation of 1st imputed and sequentially normalized gene expression data
